Ensemble Stability of Biological Networks

### 2014-09-25 paper.tex
edits

Furthermore, if it turns out that $\vec x$ is indeed a fixed point, we
may proceed to ask whether it is a dynamically stable fixed point.
Intuitively, dynamic stability means that, if one chooses the initial
conditions sufficiently close to the fixeed point, the solution will
stay close to the fixed point.  Physically, this is important because,
if a fixed point ${\vec x}_0$ is unstable, we have zero proability of
observing the solution ${\vec x}(t) = {\vec x}_0$.

@@

The spectral abscissa of the matrix $A$ is defined as
$$
\eta(A) = \max_i \{\Re(\lambda_i)\}
$$
where $\lambda_i$ are the eigenvalues of $A$. The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the spectral abscissa of $A$ is less than zero, equivalently, $\eta(A) < 0$. This is because the general solution to \ref{eq:lineardynsys} is
$$
y_i(t) = \sum_j b_{ij} e^{\lambda_j t}, \; (i=1,\ldots,n)
$$
for some matrix $B=(b_{ij})$ and thus all $\vec{y} = \vec{x} - \vec{x}^0$ decay to zero when all $\lambda_i < 0$.

Even though the determination of stability only requires examining
linearized equations of motion, it is worth noting that knowing the
stability of a fixed point can yield information about the non-linear
dynamics of a system.  For instance, in two dimensions,
Poincare-Bendixson theory implies that every limit cycle must encircle
at least one fixed point \cite{Davis1962}.  Furthermore, in the case where the cycle contains a single fixed point (such as in the Lotka-Volterra model of
conflicting populations), the stability of the fixed point will be
determined by whether the limit cycle is attractive or repulsive.

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real
coefficients and connectivity $\connectivity$ and denote the
proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may
compute the robustness of this proposition using \ref{eq:robustness}
once we have specified the proability densities $\mu$ and $\mu'$.  In
general, these will depend upon the parameters of the non-linear
system.  For the puprpose of the current investigation, we shall make
the ansatz that these are such that the distribution of entries of $A$
is (at least approximately) the uniform distribution
$\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube,
$H^\connectivity$, of edge length $r=2$, centered about the origin.
To model the perturbations, we shall resample the elements $k_1,
k_2, \ldots k_m$ of this matrix from the same uniform distribution.
Under these assumptions, the expression for robustness becomes
of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

While one can resample a fixed subset of the links, more typically we will be randomly selecting which links to resample with a uniform distribution on the links.  The result of this operation corresponds to averaging our quantity over subsets of links:
\begin{equation}
R (\hbox{avg stab}, m) =
\frac{1}{\binom{\connectivity}{m}}
\sum_{\{k_1, k_2, \ldots k_m\} \subset \{1, 2,, \ldots \connectivity\}}
R (\mathrm{stab}, k_1, \ldots k_m)
\end{equation}

In order to evaluate these integrals, one needs to describe the stable region.
This can be done in terms of conditions on the coefficients of the
characteristic polynomials associated to the systems described by matrices $A$ above \cite{Gantmacher1959}.

@@

Open systems are distinguished by having some of the variables be control variables whose values are given as specified functions of time rather than determined by the dynamics.  Given several such open systems, we may combine them to produce a larger system by setting the control variables of a subsystem equal to the dynamical variables of another system.  The composition and decomposition of open systems has algebraic structure equivalent to certain a kind of operad \cite{Vagner2014}.  Taking this notion to its logical extreme, one can dissect a system into a collection of one open system for each dynamical variable.  However, for most purposes, such a decomposition would be uninteresting and what we instead want is a decomposition into relatively self-contained modules.  One method of accomplishing this based upon the topology of the system graph is the decomposition into strongly connected components.

A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).

The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other.  Leaving the components
fixed, we may move links between nodes in a component or betwwen
components, or even add or delete links.  Some examples of these
operations are illustrated in \ref{fig:scc}B and C.  As we shall see
later, these operations leave invariant key dynamical quantities, so
we may regard them as a symmetry groupoid of our system.

@@

Using this approach the probability of stability and of robustness for all two component systems is given in \ref{tab:structstabmat}. Similarly, the analogous results for all three component systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \ref{fig:combined}B.

If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

@@

This observation suggested that graph edit distance from \ref{fig:scc}A, hierarchy, might provide a better characterization of system robustness. \ref{fig:scc}D shows system robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \ref{fig:scc}F shows system robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \ref{fig:combined}E.

@@

The fact that the most hierarchical of networks is the most robust in the case of three component systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness.  By combining our decomposition of the graph with our analysis of matrices, we will deduce a general result which relates robustness to hierarchy.

@@

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices connectivity
$\connectivity_i$ and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $R(\hbox{avg stab} (C_i), 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} (G), 1) =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i R(C_i)}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

@@

The preceding argument demonstrates the manner in which robustness is
maximized by maximizing the number of edges \emph{between} the strongly
connected components of the graph underlying system interdependencies.
And, the graph that has the latter property is that associated to the
ordering, $G_{\mathrm{tot}}$, which aslo possess the highest degree of
hierarchy. The fact that robustness is equivalent for any configuration
of strongly connected components with an equivalent number of links
between them demonstrates that it is invariant to permutations of
strongly connected components. This result does not depend at all upon
any particular details such as the probability distribution from which
the component interaction strengths are sampled because the arguments
presented here are purely topological in nature. The result that
dynamical robustness is correlated with system hierarchy therefore
applies to an even broader class of dynamical systems than the particular
random ensembles primarily studied here.

>>>>

Furthermore, if it turns out that $\vec x$ is indeed a fixed point, we
may proceed to ask whether it is a dynamically stable fixed point.
Intuitively, dynamic stability means that, if one chooses the initial
conditions sufficiently close to the fixeed point, the solution will
stay close to the fixed point.  Physically, this is important because,
if a fixed point ${\vec x}_0$ is unstable, we have zero probability of
observing the solution ${\vec x}(t) = {\vec x}_0$.

@@

The spectral abscissa of the matrix $A$ is defined as
$$
\eta(A) = \max_i \{\Re(\lambda_i)\}
$$
where $\lambda_i$ are the eigenvalues of $A$. The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the spectral abscissa of $A$ is less than zero, equivalently, $\eta(A) < 0$. This is because the general solution to \ref{eq:lineardynsys} is
$$
y_i(t) = \sum_j b_{ij} e^{\lambda_j t}, \; (i=1,\ldots,n)
$$
for some matrix $B=(b_{ij})$ and thus all $\vec{y} = \vec{x} - \vec{x}^0$ decay to zero when all $\lambda_i < 0$. This criterion can be checked equivalently in terms of conditions on the coefficients of the characteristic polynomials $\chi(A)$ associated to the systems described by matrices $A$ \cite{Gantmacher1959}.

Even though the determination of stability only requires examining
linearized equations of motion, it is worth noting that knowing the
stability of a fixed point can yield information about the non-linear
dynamics of a system.  For instance, in two dimensions,
Poincare-Bendixson theory implies that every limit cycle must encircle
at least one fixed point \cite{Davis1962}.  Furthermore, in the case where the cycle contains a single fixed point (such as in the Lotka-Volterra model of
conflicting populations), the stability of the fixed point will determine whether the limit cycle is attractive or repulsive.

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real coefficients, connectivity $\connectivity$, and denote the proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may compute the robustness of this proposition using \ref{eq:robustness} once we have specified the probability densities $\mu$ and $\mu'$.  In general, these will depend upon the parameters of the non-linear system.  For the puprpose of the current investigation, we shall make the ansatz that these are such that the distribution of entries of $A$ is (at least approximately) the uniform distribution $\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin. To model the perturbations, we resample the elements $k_1, k_2, \ldots k_m$ of this matrix from the same uniform distribution.
Under these assumptions, the expression for robustness becomes a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

While one can resample a fixed subset of the links, more typically we will be randomly selecting which links to resample with a uniform distribution on the links.  The result of this operation corresponds to averaging our quantity over subsets of links:
\begin{equation}\label{eq:robavgoverlinks}
R (\hbox{avg stab}, m) =
\frac{1}{\binom{\connectivity}{m}}
\sum_{\{k_1, k_2, \ldots k_m\} \subset \{1, 2,, \ldots \connectivity\}}
R (\mathrm{stab}, k_1, \ldots k_m)
\end{equation}
We are fundamentally interested in the relationship between \ref{eq:condprobgen} and \ref{eq:robavgoverlinks} and topological properties of systems that derive from their associated graphs. In order to investigate this further, we must characterize properties of system graphs in order to evaluate correlations of those properties with dynamical robustness.

@@

Dynamical networks of the kind described above can be viewed at the level of their interdependency graphs as being composed of more fundamental interacting subsystems. One such kind of system composition and decomposition is given by the notion of open systems, which are distinguished by having some of the variables specified as control variables whose values are given as autonomous functions of time rather than determined by the dynamics via intrinsic interactions \cite{Vagner2014}.  Given several such open systems, we may combine them to produce a larger system by setting the control variables of a subsystem equal to the dynamical variables of another system.  Taking this notion to its logical extreme, one can dissect a system into a collection of one open system for each dynamical variable.  However, this decomposition is trivial since it is equivalent to the underlying system graph and what we instead want is an intermediate decomposition into relatively self-contained modules.  One method of accomplishing this based upon the topology of the system graph is the decomposition into strongly connected components.

A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).
The decomposition of a digraph into strongly connected components corresponds to a block triangular decomposition of its adjacency matrix.  Say that the graph $G$ has strongly connected components $C_1, C_2, \ldots C_n$, which have been labelled in such a way that there are no links from vertices in component $C_i$ to component $C_j$ when $i < j$.  Label the vertices in such a way that $v_1, \ldots, v_{n_1}$ belong to $C_1$, $v_{n_1 + 1}, \ldots, v_{n_2}$ belong to $C_2$, etc.  Then, if we choose basis vectors corresponding to this labelling of the vertices, we will have $a_{ij} = 0$ whenever $i$ and $j$ correspond to different components and $i > j$.  This condition is equivalent to stating that the matrix is block triangular with blocks of size $n_1, n_2, \ldots$.

@@

\todo{add figure showing all symmetries under $\hier$}
This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other.  Leaving the components
fixed, we may move links between nodes in a component or between
components, or even add or delete links.  Some examples of these
operations are illustrated in \ref{fig:scc}B and C.  As we shall see
later, some of these operations leave invariant key dynamical quantities, so
we may regard them as a symmetry groupoid of our system.

@@

Using this approach the probability of stability and of robustness for all two component systems is given in \ref{tab:structstabmat}.

The analogous results for all three component systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \ref{fig:combined}B. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

@@

This observation suggested that graph edit distance from \ref{fig:scc}A, hierarchy, might provide a better characterization of system robustness. \ref{fig:combined}D shows system robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \ref{fig:combined}F shows system robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \ref{fig:combined}E.

@@

The fact that the most hierarchical of networks is the most robust in the case of three component systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness.  By combining our decomposition of the graph with our analysis of the robustness of random ensembles of dynamical systems, we will deduce a general result which relates robustness to hierarchy.

@@

Using these observations, we may express the robustness of a graph in
terms of the robustnesses of its strongly connected components.  If
the components $C_i$ of our graph $G$ has $v_i$ vertices, connectivity
$\connectivity_i$, and there are $\ell_{ij}$ links between components
$i$ and $j$, then there are a total of
$\sum_{i,j \in \hier(G)} \ell_{ij} + \sum_{i=1}^n \connectivity_i$
links in $G$.  If we resample a randomly chosen link in a component
$C_i$, then the probability that the system will remain stable after
resampling is $R(\hbox{avg stab} (C_i), 1)$.  If we resample a link
between components $C_i$ and $C_j$, the system will remain stable with
probability 1.  Hence, the probability that the system will remain
stable upon resampling a random link is the weighted average of these
probabilities:
\begin{align}\label{eq:sccrobustness}
R(\hbox{avg stab} (G), 1) =
\frac{\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i R(C_i)}
     {\sum\limits_{(i,j) \in \hier(G)} \ell_{ij} +
      \sum\limits_{i=1}^n \connectivity_i}.
\end{align}
For instance, if our graph is the one in \ref{fig:scc}B or C, then we have two connected components, one with two nodes, and one with one node.  From \ref{tab:structstabmat}, we know that the graph with two nodes has probability $0.25$ of being stable and and robustness $0.62$.  The graph with one node corresponds to a $1 \times 1$ matrix, so we have probability $0.5$ of stability and robustness $0.5$.  Thus, the probability of our 3-node graph being stable is $0.5 \times 0.25 = 0.125$ and its robustness is
\[
R(G) = \frac{2 + 0.5 + 4 \times 0.62}{2 + 1 + 4} = 0.714,
\]
which agrees with the value computed in \ref{tab:structstabmat} up to
sampling error.

@@

The preceding argument demonstrates the manner in which robustness is
maximized by maximizing the number of edges \emph{between} the strongly
connected components of the graph underlying system interdependencies.
And, the graph that has the latter property is that associated to the
ordering, $G_{\mathrm{tot}}$, which also possess the highest degree of
hierarchy. The fact that robustness is equivalent for any configuration
of strongly connected components with an equivalent number of links
between them demonstrates that it is invariant to permutations of
strongly connected components. This result does not depend at all upon
any particular details such as the probability distribution from which
the component interaction strengths are sampled because the arguments
presented here are purely topological in nature. The result that
dynamical robustness is correlated with system hierarchy therefore
applies to an even broader class of dynamical systems than the particular
random ensembles primarily studied here.
